---
title: "NEW EXAMPLE: County-Level Child Poverty Analysis with AI"
format:
  html:
    embed-resources: true
    self-contained: true
execute:
  echo: true
---

# NEW EXAMPLE: County-Level Child Poverty Analysis

**Estimated time:** 20 minutes  
**Packages needed:** `tidyverse`, `ggplot2`, `dplyr`, `scales`  
**Quick run:** Open in RStudio, run all chunks with Ctrl+Alt+R  

## Human+AI Workflow Overview

This exercise demonstrates a complete human+AI data analysis workflow for county-level child poverty data. You'll work step-by-step with an AI assistant to load, analyze, and visualize the data.

## Setup

```{r setup}
library(tidyverse)
library(scales)

# Read the dataset
child_poverty <- read_csv('toy-data/child-poverty-county.csv')
```

## Human+AI Workflow Steps

### Step 1: Data Exploration and Setup

**Your role:** Ask AI to help you explore and understand the dataset.

**Barebones prompt (copy-paste):**
```
Explore toy-data/child-poverty-county.csv and summarize the data structure, key variables, and any data quality issues.
```

**Expanded prompt:**
```
Load and explore the child poverty county dataset. Please provide a summary of the data structure, describe what each column represents, identify any missing values or outliers, and suggest what analysis might be appropriate for this data.
```

**Expected AI guidance:** AI should identify the GEOID, NAME, state, children_in_poverty, total_children_under18, and child_poverty_rate_pct columns and suggest county-level analysis.

```{r exploration}
# Explore the data structure
glimpse(child_poverty)
summary(child_poverty)

# Check for missing values
child_poverty %>% 
  summarise(across(everything(), ~sum(is.na(.))))

# Preview top and bottom counties by poverty rate
child_poverty %>% 
  arrange(desc(child_poverty_rate_pct)) %>% 
  head(10)
```

### Step 2: Ask AI to Compute and Validate Indicators

**Your role:** Ask AI to compute additional indicators and validate the data.

**Barebones prompt (copy-paste):**
```
Compute county child poverty rate, show top 10 counties, and validate calculations match the provided rates.
```

**Expanded prompt:**
```
Using the child poverty dataset, compute the child poverty rate from the raw counts and compare it to the provided rate column. Show the top 10 counties with highest poverty rates and verify our calculations are correct.
```

```{r compute_indicators}
# Compute poverty rate and validate
child_poverty_validated <- child_poverty %>%
  mutate(
    computed_poverty_rate = round(100 * children_in_poverty / total_children_under18, 1),
    rate_difference = abs(child_poverty_rate_pct - computed_poverty_rate)
  )

# Check validation
child_poverty_validated %>%
  filter(rate_difference > 0.1) %>%
  select(NAME, state, child_poverty_rate_pct, computed_poverty_rate, rate_difference)

# Show top 10 counties
top_poverty_counties <- child_poverty_validated %>%
  arrange(desc(child_poverty_rate_pct)) %>%
  head(10) %>%
  select(NAME, state, children_in_poverty, total_children_under18, child_poverty_rate_pct)

print(top_poverty_counties)
```

### Step 3: Ask AI to Create Visualization

**Your role:** Ask AI to create a clear, slide-ready visualization.

**Barebones prompt (copy-paste):**
```
Create a bar chart of top 10 counties by child poverty rate and save as toy-data/child_poverty_county.png.
```

**Expanded prompt:**
```
Create a horizontal bar chart showing the top 10 counties with highest child poverty rates. Make it slide-ready with clear labels, appropriate colors, and save as a PNG file suitable for presentations.
```

```{r visualization}
# Create slide-ready visualization
poverty_plot <- child_poverty %>%
  arrange(desc(child_poverty_rate_pct)) %>%
  head(10) %>%
  mutate(
    county_state = paste0(NAME, ", ", state),
    county_state = fct_reorder(county_state, child_poverty_rate_pct)
  ) %>%
  ggplot(aes(x = child_poverty_rate_pct, y = county_state)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  geom_text(aes(label = paste0(child_poverty_rate_pct, "%")), 
            hjust = -0.1, size = 3) +
  scale_x_continuous(labels = percent_format(scale = 1), 
                     expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Counties with Highest Child Poverty Rates",
    subtitle = "Top 10 counties across multiple states",
    x = "Child Poverty Rate (%)",
    y = NULL,
    caption = "Source: Toy dataset for demonstration purposes"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )

# Display plot
print(poverty_plot)

# Save plot
ggsave("toy-data/child_poverty_county.png", poverty_plot, 
       width = 10, height = 6, dpi = 300)
```

### Step 4: Ask AI to Improve and Iterate

**Your role:** Ask AI to refine the analysis and visualization based on your needs.

**Barebones prompt (copy-paste):**
```
Improve legend and caption for slides.
```

**Expanded prompt:**
```
Improve the chart for a policy presentation. Add better formatting, consider color accessibility, and provide talking points about what the data shows for child welfare advocates.
```

```{r improve_visualization}
# Improved version with better accessibility and formatting
poverty_plot_improved <- child_poverty %>%
  arrange(desc(child_poverty_rate_pct)) %>%
  head(10) %>%
  mutate(
    county_state = paste0(NAME, ", ", state),
    county_state = fct_reorder(county_state, child_poverty_rate_pct),
    # Add color coding for severity
    severity = case_when(
      child_poverty_rate_pct >= 35 ~ "Very High (35%+)",
      child_poverty_rate_pct >= 25 ~ "High (25-34%)",
      child_poverty_rate_pct >= 15 ~ "Moderate (15-24%)",
      TRUE ~ "Low (<15%)"
    )
  ) %>%
  ggplot(aes(x = child_poverty_rate_pct, y = county_state, fill = severity)) +
  geom_col(alpha = 0.9) +
  geom_text(aes(label = paste0(child_poverty_rate_pct, "%")), 
            hjust = -0.1, size = 3.5, fontface = "bold") +
  scale_fill_manual(
    values = c("Very High (35%+)" = "#d73027", 
               "High (25-34%)" = "#fc8d59",
               "Moderate (15-24%)" = "#fee08b",
               "Low (<15%)" = "#d9ef8b"),
    name = "Poverty Level"
  ) +
  scale_x_continuous(labels = percent_format(scale = 1), 
                     expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Child Poverty Crisis: Counties with Highest Rates",
    subtitle = "Over 40% of children live in poverty in some counties",
    x = "Child Poverty Rate",
    y = NULL,
    caption = "Data: Simulated county-level estimates for training purposes\nNote: Real analysis should use American Community Survey data"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12, color = "gray40"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )

print(poverty_plot_improved)

# Save improved version
ggsave("toy-data/child_poverty_county_improved.png", poverty_plot_improved, 
       width = 12, height = 8, dpi = 300)
```

## Risks & Uncertainty

When working with AI on data analysis, always consider:

### Data Currency and Quality
- **Ask AI:** "List the columns you used and any assumptions you made about the data."
- **Verify:** Check data sources, collection dates, and methodology
- **Reality check:** These are toy numbers - real analysis requires current ACS data

### Statistical Limitations  
- **Ask AI:** "What are the limitations of this analysis and what caveats should I mention?"
- **Consider:** Margin of error, sample sizes, and confidence intervals
- **Small sample warning:** Some counties may have unreliable estimates due to small populations

### AI Analysis Limitations
- **Ask AI:** "What could be wrong with this analysis approach?"
- **Verify:** Cross-check AI suggestions against domain knowledge
- **Human judgment:** AI cannot interpret policy implications or recommend actions

### Recommended Verification Steps
1. **Column verification:** Always ask AI to report which specific columns it used
2. **Calculation check:** Verify that computed rates match provided rates  
3. **Outlier investigation:** Ask AI to flag and explain any unusual values
4. **Missing data:** Confirm how AI handled any missing or zero values

## Summary

This workflow demonstrates how humans and AI can collaborate effectively on data analysis:
- **Human provides context** and domain knowledge
- **AI handles technical implementation** and suggests improvements  
- **Human validates results** and makes final decisions
- **Both iterate together** to refine analysis and presentation

The key is maintaining human oversight while leveraging AI's speed and technical capabilities.