---
title: "AI Exercises for Kids Count Data Center"
format: html
editor: visual
---

# AI Exercise Session Plan

We will work through 2 interactive exercises using AI tools. If your organization has an approved tool, feel free to use that. Otherwise, a free version of ChatGPT, Gemini, or the like will work great for illustrative purposes. As a disclaimer, these examples were aided themselves by AI tools!

This toy example deals with American Community Survey data on child poverty and showcases the usefulness of AI for catching mistakes, cleaning up messy code, and doing some preliminary data visualization.

# Exercise 1 - Documentation and Debugging

## Objective

We will start with an R function that is messy and has a bug in it. With assistance, we will clean up the code, add documentation, and fix the bug.

## Instructions

1.  Review the R code snippet below. Run it in your R environment to see what it does at baseline.

2.  There are two bugs in the current function. Draft a prompt to ask your AI tool to rewrite the function to fix the bugs. You can try without telling it the bugs at first to check what it identifies, and then provide hints if needed.

    a.  All of the margin-of-error variables (ending in '\_M') are included in the numerator alongside the estimates (ending in '\_E')
    b.  the denominator is the wrong universe - it is everyone, not just the <18 population..

3.  The function is poorly documented. Ask AI to add function docstrings to explain what the function does, its inputs, and its outputs, as well as any clarifying in-line comments that might be necessary to explain assumptions

4.  Review the refined function. Test to ensure it works as expected and note any key differences.

```{r}
library(tidyverse)
library(tidycensus)

# 1. Read in the toy dataset
county_child_pov_ar <- read_csv('toy-data/child-poverty-ar.csv')

# 2. Code to get the right list of variables in our numerator and in our denominator
year <- 2023
survey <- "acs5"
vars <- load_variables(year, survey, cache = TRUE)
  

under18_ages <- "(Under 5 years|5 years|6 to 11 years|12 to 14 years|15 years|16 and 17 years)"
  
num_vars <- vars %>%
  filter(str_detect(name, "^B17001_"),
           str_detect(label,
                      paste0("Estimate!!Total:!!Income in the past 12 months below poverty level:!!(Male|Female):!!",
                             under18_ages))) %>%
  pull(name)
 
# NOTE: den_vars calculation moved to SOLUTION section to keep the example intentionally buggy


# 2. Existing function to compute child poverty percentage by Arkansas census tract
calculate_child_pov_pct <- function(d) {
  d %>%
    transmute(
      GEOID, NAME,
      num = rowSums(across(matches(paste0("^(\\", paste(num_vars, collapse="|"), ")")))), # grab all numerator columns
      den = B17001_001E # Grab total count for the B17_001 table as denominator
    ) %>%
    mutate(child_poverty_rate_pct = 100 * num / den) %>%
    arrange(desc(child_poverty_rate_pct))
}

calculate_child_pov_pct(county_child_pov_ar)
```

---

### SOLUTION (reference)

The full solution and sample prompts live in this notebook. The corrected function below (minimal, with comments) fixes the two issues: it sums only estimate columns (excludes _M) and uses the under‑18 estimate columns as the denominator. A short sample AI prompt is provided after the code.

```{r}
# Prepare denominator variable list
den_vars <- vars %>%
  filter(str_detect(name, "^B17001_"),
         str_detect(label,
                    paste0("Estimate!!Total:!!(Income in the past 12 months (below|at or above) poverty level):!!(Male|Female):!!",
                           under18_ages))) %>%
    pull(name)

# Solution: corrected, minimal function
#' Calculate child poverty percent (under-18)
#' d: data.frame with ACS wide columns like B17001_XXXE
#' num_codes, den_codes: vectors of B17001 codes (no E/M suffix)
calculate_child_pov_pct_solution <- function(d, num_codes, den_codes) {
  num_cols <- paste0(num_codes, "E")
  den_cols <- paste0(den_codes, "E")

  num_present <- intersect(num_cols, names(d))
  den_present <- intersect(den_cols, names(d))

  d %>%
    transmute(
      GEOID, NAME,
      num = if (length(num_present) > 0) rowSums(across(all_of(num_present)), na.rm = TRUE) else 0,
      den = if (length(den_present) > 0) rowSums(across(all_of(den_present)), na.rm = TRUE) else 0
    ) %>%
    mutate(child_poverty_rate_pct = if_else(den > 0, 100 * num / den, NA_real_))
}

# Run the solution on the toy data
child_pov_pct_solution <- calculate_child_pov_pct_solution(county_child_pov_ar, num_vars, den_vars)
child_pov_pct_solution %>% slice_max(child_poverty_rate_pct, n = 6)
```

Sample AI prompt (pasteable):"

"Fix the function: use estimate (E) columns only; use <18 population as denominator; handle divide-by-zero; add docstring."

---

# Exercise 2 - Data Visualization

## Objective

We will ask AI to assist with creating data visualization and iteratively make improvements until we have what we want. 

## Instructions:

1. We will use the cleaned data from Exercise 1. If you did not complete that, you can read in this updated dataset:

```{r}
child_poverty_ar_pct <- read_csv('toy-data/child-poverty-ar-pct.csv')
```

2.  Ask your AI tool to create a heat map of child poverty rates statewide using `ggplot2.`

3.  Ask your AI tool to improve chart readability for use in a slide deck. What might we consider asking for?

4.  Ask your tool to generate a caption and short blurb to explain the narrative throughline of the chart for a non-technical audience. Can we evaluate what it generates together?---------------------------------------------------------------------

---

### SOLUTION (reference)

Below is a compact, slide-ready mapping solution. It uses a colorblind-safe palette and a county-level fallback for speed. The code saves a PNG suitable for slides.

```{r}
library(tidyverse)
library(tigris)
library(sf)
library(viridis)
library(scales)
options(tigris_use_cache = TRUE)

# Get Arkansas tract geometries
ar_tracts <- tracts(state = "AR", year = 2023, cb = TRUE)

# Join the cleaned poverty data from Exercise 1 solution
poverty_map_data <- ar_tracts %>%
  left_join(child_pov_pct_solution %>% select(GEOID, child_poverty_rate_pct, den), by = "GEOID")

# County-level aggregation (fast and slide-friendly)
county_map_data <- poverty_map_data %>%
  mutate(county_geoid = substr(GEOID, 1, 5)) %>%
  group_by(county_geoid) %>%
  summarize(child_poverty_rate_pct = weighted.mean(child_poverty_rate_pct, w = den, na.rm = TRUE), geometry = st_union(geometry), .groups = "drop")

p_county_solution <- ggplot(county_map_data) +
  geom_sf(aes(fill = child_poverty_rate_pct), color = "white", size = 0.2) +
  scale_fill_viridis_c(option = "magma", na.value = "grey90", labels = percent_format(accuracy = 1)) +
  labs(title = "Child Poverty by County — Arkansas", subtitle = "Weighted average of tract rates (denominator-weighted)", caption = "Toy data; replace with ACS for real results") +
  theme_void(base_size = 14) + theme(legend.position = "right", plot.title = element_text(face = "bold"))

# Save slide-ready image
ggsave("toy-data/child_poverty_county_slide_solution.png", p_county_solution, width = 10, height = 7, dpi = 300)

# Print preview
p_county_solution
```

Suggested prompt for visualization refinement (pasteable):

"Write R code to generate from this dataset a slide-ready county map of child poverty rates and save a PNG."

---
