---
title: "AI Exercises for Kids Count Data Center (Solutions)"
format:
  html:
    embed-resources: true
    self-contained: true
execute:
  echo: true
---

# AI Exercise Session Plan

We will work through 2 interactive exercises using AI tools. If your organization has an approved tool, feel free to use that. Otherwise, a free version of ChatGPT, Gemini, or the like will work great for illustrative purposes. As a disclaimer, these examples were aided themselves by AI tools!

This toy example deals with American Community Survey data on child poverty and showcases the usefulness of AI for catching mistakes, cleaning up messy code, and doing some preliminary data visualization.

# Exercise 1 - Documentation and Debugging

## Objective

We will start with an R function that is messy and has a bug in it. With assistance, we will clean up the code, add documentation, and fix the bug.

## Instructions

1.  Review the R code snippet below. Run it in your R environment to see what it does at baseline.

2.  There are two bugs in the current function. Draft a prompt to ask your AI tool to rewrite the function to fix the bugs. You can try without telling it the bugs at first to check what it identifies, and then provide hints if needed.

    a.  All of the margin-of-error variables (ending in '\_M') are included by default in `tidycensus`. Without logic keeping only the estimates (ending in '\_E') in the numerator, our calculations will be off base.
    b.  the denominator is the wrong universe - it is currently set to everyone, but it should only be the <18 population.

3.  The function is poorly documented. Ask AI to add function docstrings to explain what the function does, its inputs, and its outputs, as well as any clarifying in-line comments that might be necessary to explain assumptions

4.  Review the refined function. Test to ensure it works as expected and note any key differences.

```{r}
library(tidyverse)
library(tidycensus)

# 1. Read in the toy dataset
county_child_pov_ar <- read_csv('toy-data/child-poverty-ar.csv')

# 2. Code to get the right list of variables in our numerator and in our denominator
year <- 2023
survey <- "acs5"
vars <- load_variables(year, survey, cache = TRUE)
  

under18_ages <- "(Under 5 years|5 years|6 to 11 years|12 to 14 years|15 years|16 and 17 years)"
  
num_vars <- vars %>%
  filter(str_detect(name, "^B17001_"),
           str_detect(label,
                      paste0("Estimate!!Total:!!Income in the past 12 months below poverty level:!!(Male|Female):!!",
                             under18_ages))) %>%
  pull(name)
 
# NOTE: den_vars calculation moved to SOLUTION section to keep the example intentionally buggy


# 2. Existing function to compute child poverty percentage by Arkansas census tract
calculate_child_pov_pct <- function(d) {
  d %>%
    transmute(
      GEOID, NAME,
      num = rowSums(across(matches(paste0("^(\\", paste(num_vars, collapse="|"), ")")))), # grab all numerator columns
      den = B17001_001E # Grab total count for the B17_001 table as denominator
    ) %>%
    mutate(child_poverty_rate_pct = 100 * num / den) %>%
    arrange(desc(child_poverty_rate_pct))
}

calculate_child_pov_pct(county_child_pov_ar)
```

---

### SOLUTION (reference)

The full solution and sample prompts live in this notebook. The corrected function below (minimal, with comments) fixes the two issues: it sums only estimate columns (excludes _M) and uses the under‑18 estimate columns as the denominator. A short sample AI prompt is provided after the code.

```{r}
# Prepare denominator variable list
den_vars <- vars %>%
  filter(str_detect(name, "^B17001_"),
         str_detect(label,
                    paste0("Estimate!!Total:!!(Income in the past 12 months (below|at or above) poverty level):!!(Male|Female):!!",
                           under18_ages))) %>%
    pull(name)

# Solution: corrected, minimal function
#' Calculate child poverty percent (under-18)
#' d: data.frame with ACS wide columns like B17001_XXXE
#' num_codes, den_codes: vectors of B17001 codes (no E/M suffix)
calculate_child_pov_pct_solution <- function(d, num_codes, den_codes) {
  num_cols <- paste0(num_codes, "E")
  den_cols <- paste0(den_codes, "E")

  num_present <- intersect(num_cols, names(d))
  den_present <- intersect(den_cols, names(d))

  d %>%
    transmute(
      GEOID, NAME,
      num = if (length(num_present) > 0) rowSums(across(all_of(num_present)), na.rm = TRUE) else 0,
      den = if (length(den_present) > 0) rowSums(across(all_of(den_present)), na.rm = TRUE) else 0
    ) %>%
    mutate(child_poverty_rate_pct = if_else(den > 0, 100 * num / den, NA_real_))
}

# Run the solution on the toy data
child_pov_pct_solution <- calculate_child_pov_pct_solution(county_child_pov_ar, num_vars, den_vars)
child_pov_pct_solution %>% slice_max(child_poverty_rate_pct, n = 6)
```

Sample AI prompt (pasteable):"

"Fix the function: use estimate (E) columns only; use <18 population as denominator; handle divide-by-zero; add docstring."

---

# Exercise 2 - Data Visualization

## Objective

We will ask AI to assist with creating data visualization and iteratively make improvements until we have what we want. 

## Instructions:

1. We will use the cleaned data from Exercise 1. If you did not complete that, you can read in this updated dataset:

```{r}
child_poverty_ar_pct <- read_csv('toy-data/child-poverty-ar-pct.csv')
```

2.  Ask your AI tool to create a heat map of child poverty rates statewide using `ggplot2.`

3.  Ask your AI tool to improve chart readability for use in a slide deck. What might we consider asking for? A prompting tip: Providing context like why you are asking this question, specific changes you want to see, the desired format of your output, and who the target audience is can all be helpful.

4.  Ask your tool to generate a caption and short blurb to explain the narrative throughline of the chart for a non-technical audience. Can we evaluate what it generates together?---------------------------------------------------------------------

---

### SOLUTION (reference)

Below is a compact, slide-ready mapping solution. It uses a colorblind-safe palette and a county-level fallback for speed. The code saves a PNG suitable for slides.

```{r}
library(tidyverse)
library(tigris)
library(sf)
library(viridis)
library(scales)
options(tigris_use_cache = TRUE)

# Get Arkansas tract geometries
ar_tracts <- tracts(state = "AR", year = 2023, cb = TRUE)

# Join the cleaned poverty data from Exercise 1 solution
poverty_map_data <- ar_tracts %>%
  left_join(child_pov_pct_solution %>% select(GEOID, child_poverty_rate_pct, den), by = "GEOID")

# County-level aggregation (fast and slide-friendly)
county_map_data <- poverty_map_data %>%
  mutate(county_geoid = substr(GEOID, 1, 5)) %>%
  group_by(county_geoid) %>%
  summarize(child_poverty_rate_pct = weighted.mean(child_poverty_rate_pct, w = den, na.rm = TRUE), geometry = st_union(geometry), .groups = "drop")

p_county_solution <- ggplot(county_map_data) +
  geom_sf(aes(fill = child_poverty_rate_pct), color = "white", size = 0.2) +
  scale_fill_viridis_c(option = "magma", na.value = "grey90", labels = percent_format(accuracy = 1)) +
  labs(title = "Child Poverty by County — Arkansas", subtitle = "Weighted average of tract rates (denominator-weighted)", caption = "Toy data; replace with ACS for real results") +
  theme_void(base_size = 14) + theme(legend.position = "right", plot.title = element_text(face = "bold"))

# Save slide-ready image
ggsave("toy-data/child_poverty_county_slide_solution.png", p_county_solution, width = 10, height = 7, dpi = 300)

# Print preview
p_county_solution
```

Suggested prompt for visualization refinement (pasteable):

"Write R code to generate from this dataset a slide-ready county map of child poverty rates and save a PNG."

---

# NEW EXAMPLE (reference) — SOLUTION

## County-Level Child Poverty Analysis with Human+AI Workflow

**Estimated time:** 20 minutes  
**Packages needed:** R: `tidyverse`, `scales`; Python: `pandas`, `matplotlib`, `seaborn`  
**Quick run:** R: `knitr::purl("ai-exercise-new-example.qmd")` then `source("ai-exercise-new-example.R")`; Python: `jupyter nbconvert --execute ai-exercise-new-example.ipynb`  

### Overview

This example demonstrates a complete human+AI workflow for county-level child poverty analysis. Unlike the existing Arkansas tract-level example, this focuses on:
- Multi-state county-level data 
- Explicit human+AI collaboration steps
- Copy-paste prompts for efficiency
- Iterative improvement workflow

### Data Description

The toy dataset (`toy-data/child-poverty-county.csv`) contains:
- 50 counties across Alabama, Arizona, California, Colorado, Florida
- County-level child poverty counts and rates
- GEOID, NAME, state, children_in_poverty, total_children_under18, child_poverty_rate_pct

### Solution Code (R)

```{r solution_r}
library(tidyverse)
library(scales)

# Load data
child_poverty <- read_csv('toy-data/child-poverty-county.csv')

# Data validation
child_poverty_validated <- child_poverty %>%
  mutate(
    computed_poverty_rate = round(100 * children_in_poverty / total_children_under18, 1),
    rate_difference = abs(child_poverty_rate_pct - computed_poverty_rate)
  )

# Top 10 counties table
top_counties <- child_poverty %>%
  arrange(desc(child_poverty_rate_pct)) %>%
  head(10) %>%
  select(NAME, state, children_in_poverty, total_children_under18, child_poverty_rate_pct)

# Final slide-ready visualization  
poverty_plot_final <- child_poverty %>%
  arrange(desc(child_poverty_rate_pct)) %>%
  head(10) %>%
  mutate(
    county_state = paste0(NAME, ", ", state),
    county_state = fct_reorder(county_state, child_poverty_rate_pct),
    severity = case_when(
      child_poverty_rate_pct >= 35 ~ "Very High (35%+)",
      child_poverty_rate_pct >= 25 ~ "High (25-34%)",
      child_poverty_rate_pct >= 15 ~ "Moderate (15-24%)",
      TRUE ~ "Low (<15%)"
    )
  ) %>%
  ggplot(aes(x = child_poverty_rate_pct, y = county_state, fill = severity)) +
  geom_col(alpha = 0.9) +
  geom_text(aes(label = paste0(child_poverty_rate_pct, "%")), 
            hjust = -0.1, size = 3.5, fontface = "bold") +
  scale_fill_manual(
    values = c("Very High (35%+)" = "#d73027", 
               "High (25-34%)" = "#fc8d59",
               "Moderate (15-24%)" = "#fee08b",
               "Low (<15%)" = "#d9ef8b"),
    name = "Poverty Level"
  ) +
  scale_x_continuous(labels = percent_format(scale = 1), 
                     expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Child Poverty Crisis: Counties with Highest Rates",
    subtitle = "Over 40% of children live in poverty in some counties",
    x = "Child Poverty Rate",
    y = NULL,
    caption = "Data: Simulated county-level estimates for training purposes\nNote: Real analysis should use American Community Survey data"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12, color = "gray40"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )

# Save final plot
ggsave("toy-data/child_poverty_county_final.png", poverty_plot_final, 
       width = 12, height = 8, dpi = 300)

print("R solution completed successfully!")
```

### Solution Code (Python)

```{python solution_python}
import pandas as pd
import matplotlib.pyplot as plt
import numpy as np

# Load data
child_poverty = pd.read_csv('toy-data/child-poverty-county.csv')

# Data validation
child_poverty['computed_poverty_rate'] = np.round(
    100 * child_poverty['children_in_poverty'] / child_poverty['total_children_under18'], 1
)

# Top 10 counties
top_counties = child_poverty.nlargest(10, 'child_poverty_rate_pct')

# Final visualization
plt.figure(figsize=(12, 8))
plot_data = child_poverty.nlargest(10, 'child_poverty_rate_pct').copy()
plot_data['county_state'] = plot_data['NAME'] + ', ' + plot_data['state']
plot_data = plot_data.sort_values('child_poverty_rate_pct')

# Color coding
def categorize_severity(rate):
    if rate >= 35: return 'Very High (35%+)'
    elif rate >= 25: return 'High (25-34%)'
    elif rate >= 15: return 'Moderate (15-24%)'
    else: return 'Low (<15%)'

plot_data['severity'] = plot_data['child_poverty_rate_pct'].apply(categorize_severity)
color_map = {'Very High (35%+)': '#d73027', 'High (25-34%)': '#fc8d59', 
             'Moderate (15-24%)': '#fee08b', 'Low (<15%)': '#d9ef8b'}

colors = [color_map[severity] for severity in plot_data['severity']]
bars = plt.barh(plot_data['county_state'], plot_data['child_poverty_rate_pct'], 
                color=colors, alpha=0.9)

# Labels and formatting
for i, (bar, value) in enumerate(zip(bars, plot_data['child_poverty_rate_pct'])):
    plt.text(bar.get_width() + 1, bar.get_y() + bar.get_height()/2, 
             f'{value}%', ha='left', va='center', fontweight='bold')

plt.xlabel('Child Poverty Rate')
plt.title('Child Poverty Crisis: Counties with Highest Rates', fontsize=16, fontweight='bold')
plt.savefig('toy-data/child_poverty_county_python_final.png', dpi=300, bbox_inches='tight')
print("Python solution completed successfully!")
```

### Human+AI Workflow Prompts (Copy-Paste Ready)

**Step 1: Data Exploration**
```
Explore toy-data/child-poverty-county.csv and summarize the data structure, key variables, and any data quality issues.
```

**Step 2: Compute Indicators**
```
Compute county child poverty rate, show top 10 counties, and validate calculations match the provided rates.
```

**Step 3: Basic Visualization**
```
Create a bar chart of top 10 counties by child poverty rate and save as toy-data/child_poverty_county.png.
```

**Step 4: Improve Visualization**
```
Improve legend and caption for slides.
```

**Verification Prompts:**
```
List the columns you used and any assumptions.
What are the limitations of this analysis and what caveats should I mention?
What could be wrong with this analysis approach?
```

### Expected Outputs
- **Table:** Top 10 counties with poverty rates 25-44%
- **PNG files:** Both basic and improved visualizations
- **Validation:** Computed rates should match provided rates within 0.1%

### Acceptance Test
1. Both R and Python notebooks run without internet/API keys
2. Generate identical PNG outputs showing top 10 counties
3. Exercise completes in ~20 minutes
4. Human+AI workflow steps are clear and actionable
5. Risks & Uncertainty section provides specific verification guidance

---
