---
title: "AI Exercises for Kids Count Data Center"
format: html
editor: visual
---

# AI Exercise Session Plan
We will work through 2 interactive exercises using AI tools. If your organization has an approved tool, feel free to use that. Otherwise, a free version of ChatGPT, Gemini, or the like will work great for illustrative purposes. As a disclaimer, these examples were aided themselves by AI tools!

# Exercise 1 - Documentation and Debugging

## Objective
We will start with an R function that is messy and has a bug in it. With assistance, we will clean up the code, add documentation, and fix the bug.

## Instructions
1. Review the R code snippet below. Run it in your R environment to see what it does at baseline.

2. There are two bugs in the current function. Draft a prompt to ask your AI tool to rewrite the function to fix the bugs. You can try without telling it the bugs at first to check what it identifies, and then provide hints if needed.
    a. All of the margin-of-error variables (ending in '_M') are included in the numerator alongside the estimates (ending in '_E')
    b. the denominator is the wrong universe - it is everyone, not just the <18 population.

3. The function is poorly documented. Ask AI to add function docstrings to explain what the function does, its inputs, and its outputs, as well as any clarifying in-line comments that might be necessary to explain assumptions

4. Review the refined function. Test to ensure it works as expected and note any key differences.

```{r}
library(tidyverse)
library(tidycensus)

# 1. Read in the toy dataset
county_child_pov_ar <- read_csv('toy-data/child-poverty-ar.csv')

# 2. Code to get the right list of variables in our numerator and in our denominator
year <- 2023
survey <- "acs5"
vars <- load_variables(year, survey, cache = TRUE)
  

under18_ages <- "(Under 5 years|5 years|6 to 11 years|12 to 14 years|15 years|16 and 17 years)"
  
num_vars <- vars %>%
  filter(str_detect(name, "^B17001_"),
           str_detect(label,
                      paste0("Estimate!!Total:!!Income in the past 12 months below poverty level:!!(Male|Female):!!",
                             under18_ages))) %>%
  pull(name)
  
den_vars <- vars %>%
  filter(str_detect(name, "^B17001_"),
         str_detect(label,
                    paste0("Estimate!!Total:!!(Income in the past 12 months (below|at or above) poverty level):!!(Male|Female):!!",
                           under18_ages))) %>%
    pull(name)


# 2. Existing function to compute child poverty percentage by Arkansas census tract
calculate_child_pov_pct <- function(d) {
  d %>%
    transmute(
      GEOID, NAME,
      num = rowSums(across(matches(paste0("^(", paste(num_vars, collapse="|"), ")")))), # grab all numerator columns
      den = B17001_001E # Grab total count for the B17_001 table as denominator
    ) %>%
    mutate(child_poverty_rate_pct = 100 * num / den) %>%
    arrange(desc(child_poverty_rate_pct))
}

calculate_child_pov_pct(county_child_pov_ar)
```

# Exercise 2 - Data Visualization
## Objective
We will ask AI to assist with creating data visualization and iteratively make improvements until we have what we want.
## Instructions
1. We will use the cleaned data from Exercise 1. If you did not complete that, you can read in this updated dataset:
```{r}
child_poverty_ar_pct <- read_csv('toy-data/child-poverty-ar-pct.csv')
```

2. Ask your AI tool to create a bar chart of child poverty rates by county using `ggplot2`

3. Ask your AI tool to improve chart readability for use in a slide deck. What might we consider asking for?

4. Ask your tool to generate a caption and short blurb to explain the narrative throughline of the chart for a non-technical audience. Can we evaluate what it generates together?

